Extracting The Power of Dependent Types

### 2021-09-24 paper.tex
Updating ACM copyright after filling-in right transfer.
A few additions/tweaks to the parts addressing reviewer comments

%%% FIXME I am didnt update publishing rights yet, as I am waiting for
%%% the university to comeback with the payment instructions for open access.

%%% If you see 'ACMUNKNOWN' in the 'setcopyright' statement below,
%%% please first submit your publishing-rights agreement with ACM (follow link on submission page).
%%% Then please update our instructions page and copy-and-paste the NEW commands into your article.
%%% Please contact us in case of questions; allow up to 10 min for the system to propagate the information.
%%%
%%% The following is specific to GPCE '21 and the paper
%%% 'Extracting the Power of Dependent Types'
%%% by Artjoms Šinkarovs and Jesper Cockx.
%%%
\setcopyright{rightsretained} % XXX ENSURE THAT THIS VALUE IS CORRECT!!!
\acmPrice{15.00}
\acmDOI{10.1145/3486609.3487201}
\acmYear{2021}
\copyrightyear{2021}
\acmSubmissionID{splashws21gpcemain-p18-p}
\acmISBN{978-1-4503-9112-2/21/10}
\acmConference[GPCE '21]{Proceedings of the 20th ACM SIGPLAN International Conference on Generative Programming: Concepts and Experiences}{October 17--18, 2021}{Chicago, IL, USA}
\acmBooktitle{Proceedings of the 20th ACM SIGPLAN International Conference on Generative Programming: Concepts and Experiences (GPCE '21), October 17--18, 2021, Chicago, IL, USA}

@@

  Most existing programming languages provide little support to formally
  state and prove properties about programs.  Ad\-ding such capabilities is far
  from trivial, as it requires significant re-engineering of the existing
  compilers and tools.  This paper proposes a novel technique to write
  correct-by-con\-struc\-tion programs in languages without built-in verification
  capabilities, while maintaining the ability to use existing tools.
  This is achieved in three steps.  Firstly, we
  give a shallow embedding of the language (or a subset) into a
  dependently typed language.  Secondly, we write a program
  in that embedding, and we use dependent types to guarantee
  correctness properties of interest within the embedding.  Thirdly, we
  extract a program written in the original language,
  so that it can be used by the existing compilers and tools.

@@

Meanwhile, a deep embedding defines a data type to represent the AST
of the embedded language.  This requires us to define a type system
and evaluator for the language separately.  With a dependently typed
host language the AST and the type system can be defined as a single
data structure, which is typically referred to as a tagless
interpreter~\citep{PasalicTS02,Tagless}.  The benefit of deep embedding is that we get
full access to the syntactic structure of the embedded program.  This
approach works beautifully for embedded languages with simple type
systems. However, for embedded languages with dependent types, tagless
interpreters are technically
possible~\citep{CHAPMAN200921,10.1007/978-3-540-74464-1_7}, but are
\emph{extremely challenging} to define and use in practice.
For further details on why exactly this is challenging refer to the
first section of~\cite{10.1145/1863495.1863497} where the problem is meticulously
explained by~\citeauthor{10.1145/1863495.1863497}.

@@

When dealing with embedded languages that are extracted and run
by external compilers, we are fundamentally bound to have potential
semantic mismatches.  Semantics of the embedding may differ from the
semantics implemented by external compilers, as well as translation from the
embedding back to external compilers can be faulty.  Real world compilers
face exactly the same problem: how do you know that generated assembly
code is doing what you think it is doing?  Addressing such problems
requires fully verified toolchains that run on a verified hardware.
Enormous research efforts are spent towards addressing this
problem, but this is beyond the scope of this paper.

>>>>

% Updating copyright info, as per latest ACM email
% after copyright transfer.
\copyrightyear{2021}
\acmYear{2021}
\setcopyright{rightsretained}
\acmConference[GPCE '21]{Proceedings of the 20th ACM SIGPLAN International Conference on Generative Programming: Concepts and Experiences}{October 17--18, 2021}{Chicago, IL, USA}
\acmBooktitle{Proceedings of the 20th ACM SIGPLAN International Conference on Generative Programming: Concepts and Experiences (GPCE '21), October 17--18, 2021, Chicago, IL, USA}
\acmDOI{10.1145/3486609.3487201}
\acmISBN{978-1-4503-9112-2/21/10}

@@

  Most existing programming languages provide little support to formally
  state and prove properties about programs.  Ad\-ding such capabilities is far
  from trivial, as it requires significant re-engineering of the existing
  compilers and tools.  This paper proposes a novel technique to write
  correct-by-con\-struc\-tion programs in languages without built-in verification
  capabilities, while maintaining the ability to use existing tools.
  This is achieved in three steps.  Firstly, we
  give a shallow embedding of the language (or a subset) into a
  dependently typed language.  Secondly, we write a program
  in that embedding, and we use dependent types to guarantee
  correctness properties of interest within the embedding.  Thirdly, we
  extract a program written in the original language,
  so it can be used with existing compilers and tools.

@@

Meanwhile, a deep embedding defines a data type to represent the AST
of the embedded language.  This requires us to define a type system
and evaluator for the language separately.  With a dependently typed
host language the AST and the type system can be defined as a single
data structure, which is typically referred to as a tagless
interpreter~\citep{PasalicTS02,Tagless}.  The benefit of deep embedding is that we get
full access to the syntactic structure of the embedded program.  This
approach works beautifully for embedded languages with simple type
systems. However, for embedded languages with dependent types, tagless
interpreters are technically
possible~\citep{CHAPMAN200921,10.1007/978-3-540-74464-1_7}, but are
\emph{extremely} challenging to define and use in practice.
%
In particular, the mutual dependencies between the syntax for
contexts, types, and terms and the definitional equality of the object
language require large and complex inductive-inductive or
inductive-recursive to encode
\cite{10.1007/978-3-540-74464-1_7,CHAPMAN200921,10.1145/1863495.1863497}.
%
Even if we manage to give a faithful deep embedding of the language,
actually writing programs in it is still an arduous task compared
to writing programs in the meta-language directly.

@@

When dealing with embedded languages that are extracted and run
by external compilers, we are fundamentally bound to have potential
semantic mismatches.  Semantics of the embedding may differ from the
semantics implemented by external compilers, and the translation from the
embedding back to external compilers can also be faulty.  Real world compilers
face exactly the same problem: how do you know that generated assembly
code is doing what you think it is doing?  Addressing such problems
requires fully verified toolchains that run on a verified hardware.
Enormous research efforts are spent towards addressing this
problem~\cite{Leroy09,KumarMNO14}, but this is beyond the scope of this paper.
